---
title: "Group numeric variables"
format: html
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(haven) 
library(tidyverse) 
library(clean) 
library(labelled) 
library(Hmisc) 

df <- read_sav('/Users/joecrowley/R/Data/BSA Teaching Data/spss/spss25/bsa2019_poverty_open.sav', user_na = T)

```

<br>

## Benefit claimants

`NatFrEst` - number of people claiming benefits who give false info.

```{r}
attributes(df$NatFrEst) 
df %>% count(NatFrEst) %>% print(n=100)
hist(zap_missing(df$NatFrEst))
```

## ntile()

Rough buckets, ignores ties (same value can end up in differet buckets) resulting groups more or less perfectly quintiles - because does not respect ties.

```{r}
ntile(zap_missing(df$NatFrEst), n = 5) %>% freq # No labels - hard to understand
```

### Refining ntile

```{r}
# Function to add labels to ntile breaks
ntile_lab <- function(data, var, breaks = 5) { 
  
  var_to_use <- data %>% pull(var)
  
  vec <- ntile(zap_missing(var_to_use), n = breaks)
  
  breaks <- map(na.omit(unique(vec)),
                ~paste0(
                  min(zap_missing(var_to_use)[vec == .x],na.rm = T), " - ",
                  max(zap_missing(var_to_use)[vec == .x],na.rm = T)
                )
  )
  vec_unique <-  as.vector(na.omit(unique(vec)))
  names(vec_unique) <-  breaks
  vec_unique
  
  result <- vec %>% set_value_labels(.labels = vec_unique)
  
  return(result)
  
  }
```

#### Example 1

```{r}
# Test ntile_lab
ntile_lab(data = df, var = "NatFrEst") %>% as_factor() %>% fct_count()
```

**Remember**, ntile creates roughly equal groups, but as a result those groups can contain the same values, see below...

```{r}
df$chk_ntile <- ntile_lab(data = df, var = "NatFrEst") %>% as_factor()
df %>% group_by(chk_ntile) %>% summarise(
  min = min(NatFrEst, na.rm = T),
  max = max(NatFrEst, na.rm = T)
)
```

#### Example 2

```{r}
# Now with 10 breaks
ntile_lab(data = df, var = "NatFrEst", breaks = 10) %>% as_factor() %>% fct_count()
```

## cut()

From base R.

The `breaks` argument is either a vector of two or more numbers, giving the break points to cut the variable into, or a single value indicating how many roughly equal groups to split it into.

### Sequence of break points

Square bracket in output indicates the groups are capped on that side, e.g. here the right hand side includes ten in the first group.

```{r}
cut(df$NatFrEst, breaks = seq(from = 0, to = 100, by = 10)) %>% fct_count() 
```

Controlling where groups end and start is hard...

If we use the following breaks:

```{r}
seq(from = 0, to = 100, by = 10)
```

Here cut() implements those breaks so the results run from 0 up to 9, then 10 to 19, and, problematically, 100 is NA.

That is because bands close on the left (the rightmost value is excluded).

```{r}

df <- df %>% mutate(ben_cheat = cut(
    NatFrEst, 
    breaks = seq(from = 0, to = 100, by = 10), 
    include.lowest = F, 
    right = F)
  )

df %>% count(ben_cheat, NatFrEst) %>% print(n=100)
```

Now we set right = T...

-   Now breaks are 1-10, 11-20, 21-30, etc,\
-   meaning 100 is included

However, it excludes zero, because include.lowest = F.

```{r}
df <- df %>% mutate(ben_cheat = cut(
  NatFrEst, 
  breaks = seq(from = 0, to = 100, by = 10), 
  include.lowest = F, 
  right = T)
  )

df %>% count(ben_cheat, NatFrEst) %>% print(n=100)
```

So, now also set include.lowest = T

This is probably best:

-   breaks are 0-10, 11-20, 21-30,\
-   0 is included,\
-   and 100 is included.

```{r}
df <- df %>% mutate(ben_cheat = cut(
  NatFrEst, 
  breaks = seq(from = 0, to = 100, by = 10), 
  include.lowest = T, 
  right = T)
  )

df %>% count(ben_cheat, NatFrEst) %>% print(n=100)
```

With decimalised data one would need to take even more care, nonetheless the settings above work ok here.

```{r}
df %>% count(libauth) %>% print(n=1000)

# Actually the same settings as above again seem best
df <- df %>% mutate(libauth_brack = cut(
    libauth, 
    seq(1,5,by = 1), 
    include.lowest = T, 
    right = T)
  )

df %>% count(libauth_brack, libauth) %>% print(n=1000)
```

#### Handling missing values

Note, user missing values will be included if the breaks include them, see below. These would need dropping.

```{r}

# Actually the same settings as above again seem best
df <- df %>% mutate(libauth_brack = cut(
    libauth, 
    seq(-1,5,by = 1), 
    include.lowest = T, 
    right = T)
  )

df %>% count(libauth_brack, libauth) %>% print(n=1000)

```

<br>

### cut() with a preset number of groups

Using a single value to ask for a preset number of groups is particularly risky with user missing values, as below they are included by default.

```{r}

# Actually the same settings as above again seem best
df <- df %>% mutate(libauth_brack = cut(
    libauth, 
    5, 
    include.lowest = T, 
    right = T)
  )

df %>% count(libauth_brack, libauth) %>% print(n=1000)

```

More properly... with `user_na_to_na` dropping the user missing values.

*Note* the use of `dig.lab`, otherwise the break values are not clean, e.g. here the first break would be 0.996.

From documentation:

> When breaks is specified as a single number, the range of the data is divided into breaks pieces of equal length, and then the outer limits are moved away by 0.1% of the range to ensure that the extreme values both fall within the break intervals. (If x is a constant vector, equal-length intervals are created, one of which includes the single value.)

```{r}

# Actually the same settings as above again seem best
df <- df %>% mutate(libauth_brack = cut(
    user_na_to_na(libauth), 
    5, 
    include.lowest = T, 
    right = T, 
    dig.lab = 1
    )
  )

df %>% count(libauth_brack, libauth) %>% print(n=1000)

```

<br>

### Nice cut

This formats the output of cut. It assumes you'll want the breaks to close on the right and include the lowest value in the sequence of breaks.

Access function: [nice_cut](Scripts/nice_cut.R). 


```{r}

nice_cut <- function(
    variable, 
    keep_user_na = F,
    breaks,
    dig_lab = 1, 
    lab_style = 1
){
  
  if(keep_user_na == F){variable <- user_na_to_na(variable)}
  
  new_variable <- cut(
    variable, 
    breaks, 
    include.lowest = T, 
    right = T, 
    dig.lab = dig_lab
    )

  new_levels <- new_variable %>% levels()
  new_levels <- as.character(new_levels)
     
   if(lab_style == 1){
     
     new_levels <- str_replace(new_levels, "\\(|\\[", "Greater than ")
     new_levels <- str_replace(new_levels, ",", " and up to ")

    } else if (lab_style == 2) {
      
     new_levels <- str_replace(new_levels, "\\(|\\[", "")
     new_levels <- str_replace(new_levels, ",", " > & <= ")

    } else if (lab_style == 3) {
      
     new_levels <- str_replace(new_levels, "\\(|\\[", "")
     new_levels <- str_replace(new_levels, ",", "-")
     
    }

  new_levels <- str_remove(new_levels, "]")
  
  factor(
    new_variable, 
    labels = new_levels
    )

}

writeLines(
  c("nice_cut <-", deparse(nice_cut, width.cutoff = 500)),
  file.path("Scripts/", "nice_cut.R")
)

```

At most basic, it takes a vector in the `variable` argument.

```{r}

nice_cut(
  variable = df$libauth, 
  breaks = 5
) %>% fct_count()

nice_cut(
  variable = df$welfare2, 
  breaks = 5, 
  lab_style = 3
) %>% fct_count()
```

But the intention is to use it within mutate.

Example below shows max and min values within each new break.

```{r}

df %>% mutate(libauth_nice_cut = 
                nice_cut(variable = libauth,
                         breaks = 5)
              ) %>% 
  group_by(libauth_nice_cut) %>% 
  summarise(min = min(libauth, na.rm = T), max = max(libauth, na.rm = T))

```

Can also use user-defined values.

```{r}

df %>% mutate(libauth_nice_cut = 
                nice_cut(variable = libauth,
                         breaks = 1:5, 
                         lab_style = 3)
              ) %>% 
  count(libauth_nice_cut)

```

<br>

## tidyr versions to group numeric data

`cut_width()` breaks vector into chunks of a given width, here 5.

```{r}
cut_width(sample(1:100, size = 100, replace = T), width = 5) %>% fct_count() %>% print(n=21)

cut_width(zap_missing(df$NatFrEst), width = 7.5) %>% fct_count()
```

<br>

`cut_interval` n groups of equal range

```{r}
cut_interval(zap_missing(df$NatFrEst), length = 10) %>% fct_count
```

<br>

`cut_number` - n groups with approx equal observations in each

```{r}
cut_number(zap_missing(df$NatFrEst), n = 5) %>% fct_count # equal ish for sure.
```

## Quantiles

Returns a numeric vector of cut points...

```{r}
# max val is 999, it includes user NA 
quantile(df$NatFrEst)
```

<br>

```{r}
# With zap_missing, the na_rm argument is required. 
quantile(zap_missing(df$NatFrEst), na.rm=T)
```

### Can be matched with cut()

```{r}
# right = T in the cut() function gives the best result, although % not perfectly 25%. 
df <- df %>% 
  mutate(NatFrEst_quarts = 
    cut(
      NatFrEst, 
      breaks = quantile(zap_missing(df$NatFrEst), na.rm=T), 
      include.lowest = T, 
      right = T)
    )

fct_count(na.omit(df$NatFrEst_quarts), prop = T)
```

### Returning a single quantile

```{r}

# Can return a single value if you needed a 10% cut point... 
quantile(df$NatFrEst, probs = 0.1) 

# Or very detailed sequences
quantile(df$NatFrEst, probs = seq(0,1,0.1))

# Now, exclude missing and return value at which we cross 90th percentile
quantile(zap_missing(df$NatFrEst), probs = 0.9,na.rm =T)

```

#### Deriving a variable with the result

```{r}

# Cut people at the 90th percentile for highest perceptions of benefit cheating
cut_val <- quantile(zap_missing(df$NatFrEst), probs = 0.9,na.rm=T)
cut_val

df <- df %>% mutate(NatFrEst_top_10 = 
    case_when(
      is.na(zap_missing(NatFrEst)) ~ NA, 
      NatFrEst > cut_val ~ 1, 
      TRUE ~ 0), 
    NatFrEst_top_10 = factor(
        NatFrEst_top_10, 
        levels = c(0,1), 
        labels = c("Less than 75%", "More than 75%")
      )
    )

df %>% 
  count(NatFrEst_top_10) %>% 
  group_by(!is.na(NatFrEst_top_10)) %>% 
  mutate(p = n / sum(n))

```

```{r}

df %>% mutate(NatFrEst_90 = 
     cut(
       NatFrEst, 
       breaks = c(0,cut_val,100), 
       include.lowest = T, 
       right = T)) %>% 
  count(NatFrEst_90) 

```

<br>

## Weighted quantiles

```{r}
wtd.quantile(df$NatFrEst, weights = df$WtFactor) # Again, returns a numeric vector... 
```

As before can be fed into cut.

```{r}
# right = T in the cut() function gives the best result, although % not perfectly 25%. 
df <- df %>% mutate(NatFrEst_quarts_wt = 
                      cut(NatFrEst, 
                          breaks = wtd.quantile(zap_missing(df$NatFrEst), na.rm=T), 
                          include.lowest = T, 
                          right = T))

df %>% count(NatFrEst_quarts_wt)
df %>% count(NatFrEst_quarts,NatFrEst_quarts_wt) # They match, weights are weak I guess.
```

# Look at weights with quantiles...

```{r}

quantile(df$WtFactor, probs = seq(0,1,0.025)) %>% matrix(nrow = 8,ncol = 5)

quantile(df$WtFactor, probs = seq(0,1,0.025)) %>% plot

```

<br>

## Checking edits

### Random sample

```{r}

# Take a random sample...
df %>% 
  count(NatFrEst_quarts_wt, NatFrEst) %>% 
  sample_n(20) %>%
  print(n=20)

```

### Min - max values

```{r}
# Look at min and max values
df %>% group_by(NatFrEst_quarts_wt) %>% 
  summarise(min = min(NatFrEst, na.rm = T), 
            max = max(NatFrEst, na.rm = T))
```
