---
title: "labelled"
format: html
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(haven) 
library(tidyverse) 
library(clean) 
library(labelled) 
library(Hmisc) 

df <- read_sav('/Users/joecrowley/R/Data/BSA Teaching Data/spss/spss25/bsa2019_poverty_open.sav', user_na = T)

```

## The labelled package

What is useful?\
- the data dictionary tool... look_for()\
- recode skipmeal with labelled()\
- look_for()

## look_for()

```{r}

# Simplified
look_for(df, details = FALSE)

# Full output
look_for(df)

# Single variable
look_for(df, "paptype")

# Select particular variables
look_for(df, c("paptype", "RAgeCat"))

# generate_dictionary() == look_for()

```

<br>

```{r}
# results as dataframe
# all levels of variables compressed into single row. 
look_for(df, c("paptype", "RAgeCat")) %>%
  convert_list_columns_to_character()
```

```{r}
# results as LONG dataframe
# i.e. each level of variables gets its own row
look_for(df, c("paptype", "RAgeCat")) %>%
  lookfor_to_long_format() %>%
  convert_list_columns_to_character()
```

<br>

## Create labelled data

```{r}
v <- labelled(
  c(1, 2, 2, 2, 3, 9, 1, 3, 2, NA),
  c(yes = 1, no = 3, "don't know" = 8, refused = 9)
)

v

#  dplyr style for data frame
df2 <- tibble(s1 = c("M", "M", "F"), s2 = c(1, 1, 2)) %>%
  set_variable_labels(s1 = "Sex", s2 = "Question") %>%
  set_value_labels(s1 = c(Male = "M", Female = "F"), s2 = c(Yes = 1, No = 2))

df2$s2

```

<br>

## Basic labelled data edits

```{r}
# Edit a single label
val_label(v, 8)
val_label(v, 8) <- "DK"
v
```

```{r}
# val_labels removes unspecified labels
val_labels(v)
val_labels(v) <- c("Don't know" = 8)
v
```

```{r}
# Using NULL with val_labels removes all labels and haven-labelled class
val_labels(v) <- NULL
v
```

```{r}
# We can also remove value labels with...
remove_val_labels(v)
```

## More tidyverse style syntax

```{r}
# Drop all or only one label...
df2 <- df2 %>% set_value_labels(s1 = NULL) # All of a variable's labels
df2 <- df2 %>% remove_value_labels(s2 = 2) # removing one value label
val_labels(df2$s1)
val_labels(df2$s2)
```

```{r}
# Drop multiple labels at once...
df2 <- df2 %>% remove_value_labels(s2 = 8:9)
df2$s2
```

```{r}
# Within mutate, here dropping user_na and keeping variable labels. 
df2 %>% mutate(s2 = remove_labels(s2, keep_var_label = T, user_na_to_na = T)) 
```

```{r}

# set_value_labels() has the same role as val_labels, over writes all values
df2 <- df2 %>% set_value_labels(s2 = c(Yes = 1, "Don't know" = 8, Unknown = 9))
df2$s2
```

```{r}
# Whereas add_value_labels updates an existing variable (like val_label)
df2 <- df2 %>% add_value_labels(s2 = c(No = 2))
df2$s2
```

```{r}
# Reattaching labels restores haven-labelled class
val_labels(v) <- c(yes = 1, no = 3, refused = 9, "don't know" = 8)
v # By default labels appear in the order added (8 here is after 9)
```

## Apply edits across multiple variables

```{r}
# Editing multiple columns at once... 
test_df <- data.frame(v1 = 1:3, v2 = c(2, 3, 1), v3 = 3:1)
test_df 
```

All variables at once...

```{r}
val_label(test_df, 1) <- "yes"
val_labels(test_df) 
```

Specific variables, reference by name.

```{r}
val_label(test_df[, c("v1", "v3")], 2) <- "maybe"
val_label(test_df[, c("v2", "v3")], 3) <- "no"
val_labels(test_df )
```

Or use a named list.

```{r}
test_df2 <- data.frame(v1 = 1:3, v2 = c(2, 3, 1), v3 = 3:1)
val_labels(test_df2) <- list(v1 = c(no = 3, yes = 1), v2 = c(a = 1, b = 2, c = 3))
val_labels(test_df2)
```

<br>

## Sort labelled data

To sort them by their numeric values...

```{r}
sort_val_labels(v)
sort_val_labels(v, decreasing = T) # ascending is default
```

Or their labels...

```{r}
sort_val_labels(v, according_to = "l")
```

### dplyr style

```{r}
test_df2$v1
test_df2 <- test_df2 %>% mutate(v1a = sort_val_labels(v1))

# labelling order reversed
test_df2$v1a 

# but values the same 
test_df2 %>% count(v1, v1a)

```

## Handle missing data

### Tag missing values

```{r}
# Tag number 2 as a missing value (user defined style)
na_values(test_df$v1) <- (2)
na_values(test_df$v1)
test_df$v1 # Still appears as 2
```

```{r}
na_range(test_df$v2) <-  c(2, 3)

# Oddly, if missing values set with na_range, they have to be returned with na_range
na_range(test_df$v2) 

# na_values returns NULL
na_values(test_df$v2)

# And yet there are most definitely values set to user-missing
attributes(test_df$v2)
```

<br>

### Convert to NA

Missing tag to NA

```{r}
test_df$v2 <- user_na_to_na(test_df$v2)
test_df
```

### Drop missing tag

```{r}
test_df$v1 <- remove_user_na(test_df$v1)
test_df$v1
test_df
```

A more dplyr syntax style...

```{r}
test_df$v1
test_df <- test_df %>% mutate(v1 = set_na_values(v1, c(2,3)))
attributes(test_df$v1)

test_df <- test_df %>% mutate(v1 = remove_user_na(v1))
test_df$v1
attributes(test_df$v1)

```

<br>

## Drop labelled or unlabelled values

```{r}
v <- labelled(c(1, 2, 2, 2, 3, 9, 1, 3, 2, NA), c(yes = 1, maybe = 2, no = 3))
v
```

Set values with no label to NA.

```{r}
nolabel_to_na(v) # 9 is now NA, because it has no label
```

Perhaps more realistic, set only values with a label to NA...

```{r}
size <- labelled(c(1.88, 1.62, 1.78, 99, 1.91), c("not measured" = 99))
size
val_labels_to_na(size)
```

### haven-style

There are also the haven function zap_missing() and zap_labelled()

More concise in many cases.

```{r}
df %>% count(RAgeCat)
df %>% count(zap_missing(RAgeCat)) # Remove user_na only, convert to NA. 
df %>% count(zap_labels(RAgeCat)) # Remove all labels, user_na to NA by defaault
df %>% count(zap_labels(RAgeCat, user_na = T)) # Retain user_na as valid value
```

<br>

## labelled version of recode

This is an extension to the method from dplyr for labelled numeric data.

Although `recode` is called below with `dplyr`, this functionality is provided by `labelled`.

```{r}
# Recoding with the labelled package
x <- labelled(1:3, c(yes = 1, no = 2))
x

# Code below sets number 3 to be 2, and uses it's label.
dplyr::recode(x,`3` = 2L, .combine_value_labels = T)
```

Where combined values have a label, these are combined...

```{r}
x <- labelled(1:3, c(yes = 1, no = 2, maybe = 3))
x
dplyr::recode(x,`3` = 2L, .combine_value_labels = T)
```

A slightly realistic example

```{r}
# Grouping agree / disagree style scale...
x <- labelled(
  1:4,
  c(
    "strongly agree" = 1,
    "agree" = 2,
    "disagree" = 3,
    "strongly disagree" = 4
  )
)

x
```

Then, merge groups labels, .sep allows user defined break.

```{r}
dplyr::recode(
  x,
  `1` = 1L,
  `2` = 1L,
  `3` = 2L,
  `4` = 2L,
  .combine_value_labels = TRUE, 
  .sep = " or "
)
```

Here recode drops value labels, so values are combined but labels dropped. Useful where combining labels is not helpful.

```{r}
dplyr::recode(
  x,
  `1` = 1L,
  `2` = 1L,
  `3` = 2L,
  `4` = 2L,
  .keep_value_labels = F,
)
```

### tidyr style labelled recode

Create a test data frame for tidyr style recoding

```{r}
test_df <- test_df %>% set_value_labels(v1 = c(Yes = 1, No = 2, Maybe = 3))
test_df
```

Application within mutate()

```{r}
test_df$v1
test_df <- test_df %>% 
  mutate(v1a = recode(v1, 
                      `1` = 1L, 
                      `2` = 1L, 
                      .combine_value_labels = TRUE)
         )

test_df$v1a
test_df %>% count(as_factor(v1), as_factor(v1a)) # Review result
```

<br>

## Convert labelled data to factors

Application of to_factor() - convert labelled data to factors

```{r}
v <- labelled(c(1, 2, 2, 2, 3, 9, 1, 3, 2, NA), c(yes = 1, no = 3, maybe = 2))
v
```

The labelled() to_factor() function seems similar to as_factor()

```{r}
v %>% to_factor() %>% levels()
v %>% as_factor() %>% levels()
```

Options...

```{r}
v
to_factor(v) # labels as levels
to_factor(v, levels = "v") # values not labels
to_factor(v, levels ="p") # both
```

### Missing data

```{r}
to_factor(v, nolabel_to_na = T) # unalabelled values to NA
```

### Factor order

```{r}
to_factor(v, sort_levels = "n") %>% levels() # Order is the order labels were defined in
to_factor(v, sort_levels = "v") %>% levels() # Order by values 
to_factor(v, sort_levels = "l") %>% levels() # Order by labels 
```

## Convert factors to labelled data

Back to labelled (perhaps for easy edits?).

Note values now 1:4, not 1, 2, 3 and 9.

1:4 is what as.numeric would return when applied to this factor.

```{r}
f <- to_factor(v)
to_labelled(f) 
```

<br>

## Applied examples

### skipmeals

Example of skipmeals - simplify into two groups, provide my own new labels.

Could just factor() really.

```{r}
df %>% count(skipmeal)

df <- df %>% 
  mutate(skipmeal2 = case_when(
    skipmeal %in% 1:4 ~ 1, 
    skipmeal %in% 5:7 ~ 2)) %>% 
  set_value_labels(skipmeal2 = c("Less often than once a week" = 1, 
                                 "Once a week or more" = 2))

df %>% count(skipmeal2, skipmeal)
df %>% count(to_factor(skipmeal2))

```

### Edit labels

This code edits the labels on the libertarian-authoritatian scale, then drops user missing values.

```{r}
df %>% count(libauth)
val_labels(df$libauth)

df <- df %>% 
  mutate(libauth2 = add_value_labels(libauth, c("Libertarian" = 1, "Authoritarian" = 5)), 
         libauth2 = user_na_to_na(libauth2))
         
val_labels(df$libauth2)
df %>% count(libauth2) %>% print(n=100)
```

### Combine levels with recode

This has to be done exactly right.

```{r}
# Review data
freq(as_factor(df$incdiffs, levels = "both")) 
val_labels(df$incdiffs)
```

This first example shows how it can go wrong.

The 'agree' and 'strongly agree' should be combined, instead both become 'strongly agree'.

While the 'other' responses are combined into 9L - 'Not answered'.

Not very helpful

```{r}
df <- df %>% 
  mutate(incdiffs2 = recode(incdiffs, 
                      `2` = 1L, 
                      `1` = 1L, 
                      `3` = 3L,
                      `4` = 5L,
                      `5` = 5L, 
                      .default = 9L, 
                      # .combine_value_labels = T
                      ))

val_labels(df$incdiffs2) # value labels themselves unchanged

# But numeric values in the data have been recoded
df %>% count(as_factor(incdiffs2), as_factor(incdiffs))
```

Alterntaively we can set the `.combine_value_labels` argument, but this code won't run.

The issue is that df\$incdiffs contains user-defined NA values.

```{r, error=T}
attributes(df$incdiffs) # values -1, 8 and 9 are user tagged NA. 
class(df$incdiffs)

df <- df %>% 
  mutate(incdiffs2 = recode(incdiffs, 
                            `2` = 1L, 
                            `1` = 1L, 
                            `3` = 2L,
                            `4` = 3L,
                            `5` = 3L,
                            .combine_value_labels = T
  ))

```

Once we remove user defined NA, the class changes

```{r}
df$incdiffs %>% class
user_na_to_na(df$incdiffs) %>% class # no longer has haven_labelled_spss
```

<br>

This interim step (removing user NA values and the haven_labelled_spss class) needs to happen before recoding.

Now these values in the data are changed and new combined labels are also added.

```{r}
# In tidyr style, collapsing variable labels...
df <- df %>% 
  mutate(incdiffs2 = user_na_to_na(incdiffs),
         incdiffs2 = recode(incdiffs2, 
                            `1` = 1L, 
                            `2` = 1L,
                            `3` = 2L,
                            `4` = 3L,
                            `5` = 3L,
                            .combine_value_labels = T)) 

df %>% 
  select(incdiffs, incdiffs2) %>% 
  mutate_all(~to_factor(.x, levels ="p")) %>% 
  count(incdiffs, incdiffs2)
```
