---
title: "Missing data"
format: html
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(here)
here::i_am("Missing-data.qmd")
library(haven) 
library(tidyverse) 
library(clean) 
library(labelled) 
library(Hmisc) 

df <- read_sav(here::here("data", "synthetic_bsa2019_sav.sav"), user_na = TRUE)

```

## drop_na

I like this function, it's not complicated or anything but it works nicely.

It retains `complete` rows. 

```{r}

df %>% nrow 
df %>% drop_na %>% nrow # By default, user NA values are not dropped
df %>% remove_user_na() %>% drop_na() %>% nrow() # Valid rows the same after these values have their missing tag dropped
df %>% user_na_to_na() %>% drop_na() %>% nrow() # But increases hugely when they are set to `NA`.

df %>% count(RAgeCat)
df %>% 
  mutate(RAgeCat = user_na_to_na(RAgeCat)) %>%
  drop_na(RAgeCat) %>% 
  nrow

```

## coalesce

From help: 

Given a set of vectors, coalesce() finds the first non-missing value at each position. 

It's inspired by the SQL COALESCE function which does the same thing for SQL NULLs.

```{r}
# BSA is not great for this... so creating a test dataset
coalesce_data <- 
  data_frame(
  v1 = sample(c(1:10, NA), 100, replace = T), 
  v2 = sample(c(1:10, NA), 100, replace = T),
  v3 = sample(c(1:10, NA), 100, replace = T)
)

coalesce_data 
```

Replace NA with the first non-missing value, going in order...

```{r}
coalesce_data %>% 
  mutate(
    non_miss = coalesce(v1,v2,v3),
    non_miss_reverse = coalesce(v3,v2,v1)
  ) %>% 
  print(n=30)
```

A more tidy-select sort of approach

```{r}
coalesce_data %>% 
  mutate(
    non_miss = reduce(across(everything()), coalesce)
    ) %>% 
  print(n=30)
```

