---
title: "Numeric variables"
format: html
editor: visual
---

```{r}
#| echo: false
#| message: false
#| warnings: false
library(haven) 
library(tidyverse) 
library(clean) 
library(labelled) 
library(Hmisc) 

df <- read_sav('/Users/joecrowley/R/Data/BSA Teaching Data/spss/spss25/bsa2019_poverty_open.sav', user_na = T)

```

<br>

## Mean of single variables

```{r}
df <- df %>% mutate(libauth_av = mean(libauth, na.rm = T), 
                    leftrigh_av = mean(leftrigh, na.rm = T))

df %>% count(libauth_av)
df %>% count(leftrigh_av)
```

<br>

## Mean of multiple variables at once (each mean is separate)

### A purrr-style formula

```{r}

df <- df %>% mutate(
  across(
    c(welfare2:libauth), 
    ~mean(.x, na.rm = T), 
    .names = "{.col}_mean")
  )

df %>% count(libauth_mean)
df %>% count(leftrigh_mean)
df %>% count(welfare2_mean)

```

<br>

## Calculate multiple statistics per variable

### Requires a named list of functions

Adds a column with the mean of `welfare2` and the standard deviation.

```{r}

df %>% select(welfare2) %>% 
  mutate(
    across(
      all_of("welfare2"), 
      # First part is name for stat column, second is the statistic function
      list(var_mean = mean, sd = sd), 
      .names = "{.col}_{.fn}")
    )

```

## Selecting variables to edit

### Selecting variables based on type

Can select variables to be edited based on variable type as well.

```{r}

df %>% select(Sserial, WtFactor, welfare2, libauth) %>%
  mutate(
    across(
      where(is.double) & !c(Sserial),
      # Otherwise, name comes out as WtFactor_1 
      list(rounded = round), 
      .names = "{.col}_{.fn}"
      )
    ) %>% 
  glimpse

```

<br>

### Selecting based on location

```{r}
df %>% mutate(across(2, round))
```

<br>

### Mean across multiple variables

Note variables calculated column-wise.

DOES NOT WORK - this takes the average only of the first variable

```{r}
df <- df %>% mutate(av_ati = mean(libauth, leftrigh, na.rm = T))
df %>% count(libauth, leftrigh, av_ati)
```

Mean of both variables for whole sample

```{r}
df <- df %>% mutate(av_ati = mean(c(libauth, leftrigh), na.rm = T))
df %>% count(av_ati)
```

Mean of both variables allowing tidy select - use c_across

```{r}
df %>% mutate(av_ati2 = mean(c_across(all_of(c("libauth", "leftrigh"))), na.rm = T)) %>% count(av_ati2)
```

<br>

## Row or group-wise operations

### Row-wise variable summaries \[SLOW\]

```{r}
df %>% 
  rowwise() %>%
  mutate(av_ati2 = mean(c_across(all_of(c("libauth", "leftrigh"))), na.rm = T)) %>% 
  count(libauth, leftrigh, av_ati2)
```

<br>

### Quicker row-wise functions...

```{r}
df <- df %>% mutate(
  av_ati3 = rowMeans(across(c(libauth, leftrigh)), na.rm = T)
)

df %>% select(av_ati3, libauth, leftrigh) %>% 
  sample_n(30) %>% print(n=1000)
```

<br>

#### Another example

```{r}
set.seed(123)  # for reproducibility

num_df <- data.frame(
  var1 = sample(1:100, 50, replace = T),
  var2 = sample(1:45, 50, replace = T),
  var3 = sample(1:20, 50, replace = T),
  var4 = sample(1:75, 50, replace = T),
  var5 = sample(1:50, 50, replace = T)
)

num_df <- num_df %>% mutate(
  
  # All columns
  row_total = rowSums(across(everything()), na.rm = T), 
  
  # Identify column numbers
  rows_2_4 = rowSums(across(c(2,4)), na.rm = T), 
  
  # Identfy variable with ends_with
  ends_with_3_5 = rowSums(across(ends_with(c("3","5")))),
  
  # With character vector
  `1_and_2` = rowSums(across(all_of(c("var1", "var2"))))
  
)

num_df %>% print()
```

<br>

### Group wise variable summaries

```{r}
df %>% 
  group_by(RSex) %>%
  mutate(av_ati2 = mean(c_across(all_of(c("libauth", "leftrigh"))), na.rm = T)) %>% 
  count(av_ati2) # Remember, without ungroup output is still grouped... 
```
